Making Sequential Programs Deterministic

### 2019-03-20 design.tex
polymorphism

%collections type parameters - invariant
Collection types are invariant. So, if a \<List> x is declared as \<@A List><\<@A String>> \<x;> and another \<List y> is declared
as  \<@A List><\<@B String>> \<y;>, writing \<x = y> would result in an error.

@@

\subsection{Additional typing rules for collection elements}
Lists, set and maps - element types have to be subtypes of collection type 
Explain iterator and extracting elements of an ordernondet collection here
Examples of valid and invalid types

\subsection{Avoid aliasing unsoundness}
Give an example of unsoundness 
Recall CLIMB-to-top rule and defaulting choice for local collections (to avoid too many warnings)
Required user annotations for collections and arrays

\subsection{Special cases of polymorphism}
Explain polymorphism first
Explain modifiers to Polymorphic annotations
Poly up and down for ordernondetermistic types
Poly use for avoiding polluting det/ordernondet collections
Subtype hierarchy among all variants of Poly
Possibility of type refinement?

\subsection{Discussion on defaulting rules}
Det - default, but @Poly is the default for parameters and return types
Choice of defaults on class type, constructors and method params and return types

@@

>>>>

%collections type parameters - invariant
Collection types are invariant. So, if a \<List> x is declared as \codeid{@A List<@A String> x;} and another \<List y> is declared
as  \codeid{@A List<@B String> y;}, writing \<x = y> would result in an error.

@@

\subsection{Additional typing rules for collection elements}
The (determinism) type of a Collection or Iterator must be a supertype or equal to
the type of the type parameter (see Figure~\ref{fig-determinism-collections}).
If the Collection is \<@NonDet>, then the type parameter may not be
\<@Det> or \<@OrderNonDet>. Although such types could exist, they are
disallowed to prevent the following bug:

\begin{verbatim}
public static void add(@NonDet List<@Det String> list, @NonDet int index, @Det String s) {
    list.add(index, s);
}

public static void f(@Det List<@Det String> list, @NonDet int index, @Det String s) {
    add(list, index, s);
}
\end{verbatim}

The possibility of mutation allows us to add to the \<@Det List> at a
\<@NonDet> index, which is unsound.

Some examples of valid types are:
\begin{itemize}
	\item \codeid{@NonDet\ \ \ \ \ \ List<@NonDet\ \ \ \ \ \ Integer>}
	\item \codeid{@Det\ \ \ \ \ \ \ \ \ List<@Det\ \ \ \ \ \ \ \ \ Integer>}
	\item \codeid{@OrderNonDet Set <@OrderNonDet Set<...>\relax >}
	\item \codeid{@OrderNonDet Set <@Det String>}
	\item \codeid{@Det\ \ \ \ \ \ \ \ \ Map <@Det String, @Det Integer>}
	\item \codeid{@NonDet\ \ \ \ \ \ Set <T extends @NonDet Object>}
\end{itemize}

These types are invalid:
\begin{itemize}
	\item \codeid{@OrderNonDet Set <@NonDet\ \ \ \ \ \ Integer>}
	\item \codeid{@NonDet\ \ \ \ \ \ List<@Det\ \ \ \ \ \ \ \ \ Integer>}
	\item \codeid{@NonDet\ \ \ \ \ \ List<@OrderNonDet Set<...>\relax >}
	\item \codeid{@Det\ \ \ \ \ \ \ \ \ List<@NonDet\ \ \ \ \ \ Integer>}
	\item \codeid{@Det\ \ \ \ \ \ \ \ \ List<@OrderNonDet Set<...>\relax >}
	\item \codeid{@Det\ \ \ \ \ \ \ \ \ Map <@Det String, @OrderNonDet List<@Det String>>}
	\item \codeid{@Det\ \ \ \ \ \ \ \ \ List<T extends @NonDet Object>}
	\item \codeid{@NonDet\ \ \ \ \ \ Set\ <T extends @Det Object>}
\end{itemize}

Similarly, the (determinism) type of an array must be a supertype or equal to
the type of the component type (see Figure~\ref{fig-determinism-collections}).
As with collections, \codeid{@NonDet} arrays of \codeid{@Det} or \codeid{@OrderNonDet}
elements are not allowed.

For example, these types are valid:
\begin{itemize}
	\item \codeid{@NonDet int @NonDet \ \ \ \ []}
	\item \codeid{@Det \ \ \ int @Det \ \ \ \ \ \ \ []}
	\item \codeid{@Det \ \ \ int @OrderNonDet[] @OrderNonDet[]}
\end{itemize}

These types are invalid:
\begin{itemize}
	\item \codeid{@NonDet int @OrderNonDet[]}
	\item \codeid{@NonDet int @Det \ \ \ \ \ \ \ []}
	\item \codeid{@Det \ \ \ int @NonDet \ \ \ \ []}
\end{itemize}

\begin{figure}
	\centering
	\begin{tabular}{|l|l|l|l|l|}
		\cline{3-5}
		\multicolumn{2}{c|}{~}  &  \multicolumn{3}{c|}{Type Argument (or array component type)} \\ \cline{3-5}
		\multicolumn{2}{c|}{~}  & NonDet     & OrderNonDet & Det \\ \hline
		& NonDet      &   valid    &  invalid    & invalid  \\ \cline{2-5}
		Collection (or array)   & OrderNonDet &   invalid  &  valid      & valid  \\ \cline{2-5}
		& Det         &   invalid  &  invalid    & valid      \\ \hline
	\end{tabular}
	\caption{Valid Collection (and array) declarations.  The Collection's (or array's) type qualifier
		must be a supertype or equal to the type argument (or array component type).}
	\label{fig-determinism-collections}
\end{figure}

A collection (\<List, Set, or a Map>) annotated as \<@OrderNonDet> has the following properties:
\begin{enumerate}
	\item The individual elements extracted by iterating over the collection get the type \<@NonDet>.
	\item Accessing properties sich as \<size()> or querying if the collection \<isEmpty()> will return types
	that are annotated as \<@Det> as they do not depend on the iteration order. 
\end{enumerate}

A user defined type may be annotated as \<@OrderNonDet> if and only if it
is a subtype of Collection or Iterator.
For example, if a user defines a type as
\begin{verbatim}
public class TestUserCollection<E> extends ArrayList<E> {...}
\end{verbatim}
Writing \codeid{@OrderNonDet TestUserCollection<@Det Integer>} is valid.\\
Writing \codeid{@OrderNonDet TestUserCollection<@NonDet Integer>} is invalid
because the type of the type parameter (\codeid{@NonDet}) is not a subtype
of the user defined Collection type (\codeid{@OrderNonDet}).

User defined types that are not a subtype of Collection or Iterator
may not be annotated as \codeid{@OrderNonDet}.

\subsection{Polymorphic annotations and special cases}
The underlying checker-framework on which we build our determinism checker has support for
polymorphic annotations. When a user writes a polymorphic annotation on a method signature or a type parameter,
it indicates that it could resolve to any type in the type system depending on how it is used.
In the determinism checker, we define a polymorphic annotation \<@PolyDet>.
One of the most common locations for a polymorphic annotation is at a method signature.
Consider the following method declaration annotated with polymorphic annotations
\begin{verbatim}
@PolyDet int foo (@PolyDet String param1, @PolyDet boolean param2) { }
\end{verbatim}
This indicates that method \<foo> can be called with arguments having any of the \codeid{@Det or @NonDet} type annotations 
(\<@OrderNonDet> is not allowed because it is invalid on primitive types). \<@PolyDet> resolves to the least upper bound of
the actual types on arguments. For instance, if this method is called as \codeid{foo(@Det String arg1, @NonDet boolean arg2)}, the 
checker framework resolves the method declaration as \codeid{@NonDet int foo(@NonDet String param1, @NonDet boolean param2)}
causing the return type to have the type annotation \<@NonDet>.

Recall that the determinism checker does not permit \<@OrderNonDet> on a type that isn't a collection, iterator, or an array.
As a consequence, resolution of polymorphic type annotations at method signatures could have undesirable effects.
Suppose a method declaration is annotated as \codeid{@PolyDet int size(@PolyDet List<T> list)}. Invoking this method
as \codeid{size(@OrderNonDet List<@Det String>)} will result in \<@PolyDet> getting resolved to \<@OrderNonDet> which is invalid
on the return type \<int>. We circumvent this scenario by introducing two variants to the \<@PolyDet> annotation, \<@PolyDet("up")>
and \<@PolyDet("down")>. The semantics of these variants are as follows:
\begin{itemize}
	\item \<@PolyDet("up")>: Same as \<@PolyDet> except when it resolves to \<@OrderNonDet>, it is replaced with \<@NonDet>.
	\item \<@PolyDet("down")>: Same as \<@PolyDet> except when it resolves to \<@OrderNonDet>, it is replaced with \<@Det>.
\end{itemize} 
The modifiers "up" and "down" are indicative of the subtyping relationship of \<@OrderNonDet> with \<@NonDet> and \<@Det> respectively
in the determinism type hierarchy.
Annotating \<size()> as \codeid{@PolyDet("down") int size(@PolyDet List<T> list)} will only allow the following type resolutions:
\begin{itemize}
	\item \codeid{@Det int size(@Det List<T> list)}
	\item \codeid{@Det int size(@OrderNonDet List<T> list)}
	\item \codeid{@NonDet int size(@NonDet List<T> list)}
\end{itemize}

In order to provide soundness guarantees, we must ensure that a collection that is
declared as \<@Det> is not side-effected in non-deterministic ways. Consider the following annotation 
on the List add method: \codeid{public void add(@PolyDet ArrayList<E> this, @PolyDet int index, E element)}.
Suppose a client calls this method as follows:
\begin{verbatim}
@Det ArrayList<@Det String> list;
@NonDet int random;
@Det String str;
list.add(random, str);
\end{verbatim}
This will result in the instantion \codeid{public void add(@NonDet ArrayList<E> this, @NonDet int index, E element)} and
the method invocation will type check. This is problematic because even though \<list> was declared to be \<@Det>,
the type checker allows the addition of an element at a \<@NonDet> location violating our determinism guarantees.
We prevent this behavior by introducing another valiant of \<@PolyDet>, \<@PolyDet("use")> that does not affect
type instantiations. This is especially useful in preventing methods from non-deterministically modifying the state
of a deterministic receiver. To have this effect, the library method \<ArrayList.add> is annotated as:
\codeid{public void add(@PolyDet ArrayList<E> this, @PolyDet("use") int index, E element)}.
Since \<@PolyDet("use")> has no effect on type instantiations, calling this method on a \<@Det List>
will result in the method declaration being resolved as \codeid{public void add(@Det ArrayList<E> this, @Det int index, E element)}
irrespective of the type of \<index>, thereby preventing undesired side-effects.

\subsection{Dataflow analysis}

\subsection{Type Refinement}

\subsection{Discussion on defaulting rules}
Det - default, but @Poly is the default for parameters and return types
Choice of defaults on class type, constructors and method params and return types
Defaults on hashsets, hashmap, etc

\subsection{Avoid aliasing unsoundness}
Recall CLIMB-to-top rule and defaulting choice for local collections (to avoid too many warnings)
Required user annotations for collections and arrays

@@

\subsection{Maps discussion}
@Det --> keySet iteration is @Det
get(): can return o-nd collections

\subsection{About generics and upper bounds?}
